---
title: "DA5020.Practium1.Luo.Rmd"
author: "Xin Lin, Xinxin Luo, William Morrison"
output:
  html_document
---
```{r}
library(tidyverse)
library(ggplot2)
library(dplyr)
library(readr)
library(XML)
library(RCurl)
library(lubridate)
library(knitr)
```

## Question 1
Load the XML file into a browser or text editing tool and inspect it. Explore the data set as you see fit and that allows you to get a sense of the data and get comfortable with it.
```{r}
#Loading the dataset
xmldata<-getURL("https://da5020.weebly.com/uploads/8/6/5/9/8659576/pubmedsample.jun18.xml")
xmldf<-xmlToDataFrame(xmldata)

# Convert data into tree
parsed_data<-xmlTreeParse(xmldata, useInternalNodes = TRUE)
medRoot<-xmlRoot(parsed_data)
```

## Question 2
Load the data into R and create two linked tibbles: one for publication and one for journal. 
Use ISSN as the key to link them. Only load the following information into the publication tibble: PMID (primary key for publication), ISSN and publication year (foreign key for journal), date completed (as one date field), date revised (as one date field), number of authors (derived/calculated field from the authors), publication type, title of article.
Load this information into the journal tibble: ISSN (primary key), medium (from CitedMedium attribute), publication year (primary key), publication season, language, and journal title. In cases where there are multiple languages for a publication, pick the first language. Same for publication type: pick the first one. 
The primary key for journal is (ISSN, publication year). Also, exclude any journals that do not have an ISSN as the primary key cannot be empty.
```{r}
# Parsing information for tibbles
# PMID number
parsePMID <- function(rNode) {
  pmID <- xmlValue(rNode[["MedlineCitation"]][["PMID"]][["text"]])
  return(pmID)
}

# ISSN number
parseISSN <- function(rNode) {
  issn <- xmlValue(rNode[["MedlineCitation"]][["Article"]][["Journal"]][["ISSN"]][["text"]])
  return(issn)
}

# Publication year
parsePubYear <- function(rNode) {
  pubYear <- xmlValue(rNode[["MedlineCitation"]][["Article"]][["Journal"]][["JournalIssue"]][["PubDate"]][["Year"]][["text"]])
  return(pubYear)
}

# Date completed
parseCompDate <- function(rNode) {
  compDate <- xmlValue(rNode[["MedlineCitation"]][["DateCompleted"]])
  return(compDate)
}

# Date revised
parseRevDate <- function(rNode) {
  revDate <- xmlValue(rNode[["MedlineCitation"]][["DateRevised"]])
  return(revDate)
}

# Number of authors
parseNumAuth <- function(rNode) {
  authors <- xmlSize(rNode[["MedlineCitation"]][["Article"]][["AuthorList"]])
  numAuthors <- as.numeric(authors)
  return(numAuthors)
}

# Publication Type
parsePubType <- function(rNode) {
  pubType <- xmlValue(rNode[["MedlineCitation"]][["Article"]][["PublicationTypeList"]][["PublicationType"]])
  return(pubType)
}

# Article Title
parseArtTitle <- function(rNode) {
  title <- xmlValue(rNode[["MedlineCitation"]][["Article"]][["ArticleTitle"]][["text"]])
  return(title)
}

# Cited Medium
Medium <- xpathSApply(parsed_data, "//PubmedArticle/MedlineCitation", function(x) {
  if (xpathSApply(x, "boolean(./Article/Journal/JournalIssue)")) {
    xpathSApply(x, "./Article/Journal/JournalIssue", xmlGetAttr, "CitedMedium")
  }
  else {
    NA
  }
})

# Publication Season
parsePubSeason <- function(rNode) {
  pubSeason <- xmlValue(rNode[["MedlineCitation"]][["Article"]][["Journal"]][["JournalIssue"]][["PubDate"]][["Season"]][["text"]])
  return(pubSeason)
}

# Lanuage
parseLang <- function(rNode) {
  lang <- xmlValue(rNode[["MedlineCitation"]][["Article"]][["Language"]])
  return(lang)
}

# Journal Title
parseJourTitle <- function(rNode) {
  title <- xmlValue(rNode[["MedlineCitation"]][["Article"]][["Journal"]][["Title"]][["text"]])
  return(title)
}

# Construct tibbles####
# Initialize tibbles
publication <- tibble(pmID = character(),
                      issn = character(),
                      pubYear = character(),
                      dateComp = date(),
                      dateRevi = date(),
                      authors = numeric(),
                      pubType = character(),
                      title = character())
journal <- tibble(issn = character(),
                  medium = character(),
                  pubYear = date(),
                  pubSeason = character(),
                  lang = character(),
                  jourTitle = character())

# Loop through the tree to get information
numNodes <- xmlSize(medRoot)

for (i in 1:numNodes)
{
  # turn off parse warnings
  options(warn = -1)
  
  # work on the "ith" listing
  currListing <- medRoot[[i]]
  
  # parse auction ID and add as PK for listing tibble
  pk <- parsePMID(currListing)
  # ensure that PK is unique, i.e., is not already in listing tibble
  if (nrow(publication[publication$pmID == pk,]) == 1)
  {
    # it already exists, so somehow the listing is already there so skip everything below
    # and continue with the next iteration of the loop
    print (paste("pmID already exists:", pk), quote = FALSE)
    next
  }
  
  # add data to publication tibble
  publication[i,]$pmID <- pk
  publication[i,]$issn <- parseISSN(currListing)
  publication[i,]$pubYear <- parsePubYear(currListing)
  publication[i,]$dateComp <- parseCompDate(currListing)
  publication[i,]$dateRevi <- parseRevDate(currListing)
  publication[i,]$authors <- parseNumAuth(currListing)
  publication[i,]$pubType <- parsePubType(currListing)
  publication[i,]$title <- parseArtTitle(currListing)
  
  # add data to journal tibble
  journal[i,]$issn <- parseISSN(currListing)
  journal[i,]$medium <- Medium[i]
  journal[i,]$pubYear <- parsePubYear(currListing)
  journal[i,]$pubSeason <- parsePubSeason(currListing)
  journal[i,]$lang <- parseLang(currListing)
  journal[i,]$jourTitle <- parseJourTitle(currListing)
}

# Change the class to "Dates"
publication$dateComp <- ymd(publication$dateComp, tz = "UTC")
publication$dateRevi <- ymd(publication$dateRevi, tz = "UTC")

# Extract valid records
journal <- filter(journal, issn != "NA", pubYear != "NA")
publication <- filter(publication, issn != "NA", pubYear != "NA")
journal <- unique(journal)

# Display the publication and journal tibbles
kable(publication)
kable(journal)
```

## Question 3
Create a line graph of the number of publications per year from 2000 to 2015.
```{r eval=TRUE}
# find number of publications per year from 2000 to 2015
pubsPerYear <- publication %>%
  filter(pubYear >= 2000, pubYear <= 2015) %>%
  count(pubYear)

# Plot the results  
ggplot(data = pubsPerYear, aes(x = pubYear, y = n)) +
  geom_point()
plot(pubsPerYear$pubYear, type = "o", xlab = "Number of Publications", ylab = "Publication Year")
```

## Question 4
Find the articles that had fewer than three authors and list the article, journal, and publication date.
```{r eval=TRUE}
# find artciles with 1-3 authors
threeOrLessAuthors <- publication %>%
  inner_join(journal, by = c("issn", "pubYear")) %>%
  filter(authors < 4, authors > 0) %>% 
  select(title, jourTitle, dateComp)

# find articles with 0 author
zeroAuthors <- publication %>%
  inner_join(journal, by = c("issn", "pubYear")) %>%
  filter(authors == 0) %>%
  select(title, jourTitle, dateComp)

# print the results
kable(threeOrLessAuthors)
kable(zeroAuthors)
```
In results, there are 78 articles with 1 to 3 authors, 92 with "0" to 3 authors.

## Question 5
Find the average number of authors for articles. Display a single number.
```{r eval=TRUE}
mean(publication$authors)
```
The average number of authors for articles is almost 10.

## Question 6
What is the average time period (in days) between date completed and date revised. Display the time elapsed in days. Only consider cases where the difference is a positive number.
```{r eval=TRUE}
# Select articles that has positive difference between date completed and date revised 
timePeriod <- publication %>%
  filter(dateComp < dateRevi) %>%
  select(issn,title, dateComp, dateRevi)

# Initialize the parameters
timeRow <- nrow(timePeriod)
timeInterval <- NA * timeRow
timePeriod["timeElapsed"] <- NA

# Loop through all rows
for (i in 1:timeRow) {
  timeInterval[i] <- interval(timePeriod$dateComp[i], timePeriod$dateRevi[i])
  timePeriod$timeElapsed[i] <- as.duration(timeInterval[i]) / ddays(1)
}

# display the table
kable(timePeriod)

# average time period
mean(timePeriod$timeElapsed)
```
The average time period between date completed and date revised is 3796 days.

## Question 7
Which articles published in PubMed were not written in English? Only consider the first language of publication.
```{r eval=TRUE}
nonEng <- journal %>%
  filter(lang != "eng") %>%
  left_join(publication, by = c("issn", "pubYear")) %>%
  select(jourTitle, title, lang)
kable(nonEng)
```
The table above listed out the articles published in PubMed were not written in English.

## Question 8
Using the XML data (not the tibbles created above), find the articles containing any of the words "drug resistance" or "virus"  in any capitalization in the title. Note that drug resistance could be spelled as "drug resistance" or "drug-resistance" or "drug resistant" or "drug resistent" -- use regular expressions to deal with the variations.
```{r eval=TRUE}
med01 <- readLines("http://da5020.weebly.com/uploads/8/6/5/9/8659576/pubmedsample.jun18.xml")

drug <- regexpr("<ArticleTitle>(.*)[D|d]rug(.*)[R|r]esist(.*?)</ArticleTitle>", med01)#regular expression that finds titles with variations on the term "drug resistance".
d <- regmatches(med01, drug)#saves matched terms as a vector.
(drugTitles <- gsub("<ArticleTitle>|</ArticleTitle>", "", d))#removes the "<ArticleTitle>" tags so only the title is printed.

virus <- regexpr("<ArticleTitle>(.*)[V|v]irus(.*)</ArticleTitle>", med01)#regular expression that finds titles with variations on the term "virus".
v <- regmatches(med01, virus)#saves matched terms as a vector.
(virusTitles <- gsub("<ArticleTitle>|</ArticleTitle>", "", v))#removes the "<ArticleTitle>" tags so only the title is printed.
```
The listed 3 articles containing any of the words "drug resistance" or "virus".
